k-plus Anticlustering: An Improved k-means Criterion for Maximizing Between-Group Similarity

### 2024-03-05 paper.Rmd
abstract

    When assigning stimuli to experimental conditions, researchers strive for high similarity between stimulus sets and high pairwise dissimilarity of stimuli within sets. Because stimulus assignment is usually inadequate when done manually, automated algorithmic approaches are preferred. Brusco et al. (2020) proposed a bicriterion algorithm (BILS) to obtain stimulus partitionings that simultaneously satisfy two anticlustering criteria: the diversity, corresponding to between-set similarity, and the dispersion, corresponding to pairwise within-set dissimilarity. As a heuristic, however, the BILS does not necessarily identify globally optimal criterion values. We present a novel exact algorithm that returns (multiple) globally optimal partitions with regard to the dispersion. Despite its theoretical computational hardness, our method scales to rather large data sets and can be applied to solve most real-world use cases. In a simulation study, we evaluate several options to combine our optimal approach with the BILS heuristic, which -- due to its bicriterion approach -- naturally lends itself to providing high diversity on top of an optimal dispersion. We also provide two further extensions to the BILS. First, we allow that the dispersion and the diversity can be optimized on the basis of different dissimilarity matrices. Second, we allow that other anticlustering objectives of between-group similarity -- such as the k-means or k-plus objectives -- can be optimized instead of only the diversity. The original BILS as well as our extensions are freely available via the free and open source R package `anticlust`. Practical examples are included to illustrate their application.

>>>>

    Numerous applications in psychological research require that a data set is partitioned via the maximization---instead of the more common minimization---of a clustering criterion to obtain high similarity between groups and heterogeneity within groups. Such anticlustering may consider criteria of overall between-group similarity and/or pairwise within-group dissimilarity. Brusco et al. (2020) proposed a bicriterion algorithm (BILS) to obtain partitionings that simultaneously maximize the diversity, corresponding to between-group similarity, and the dispersion, corresponding to pairwise within-group dissimilarity. As a heuristic, however, the BILS does not necessarily identify globally optimal criterion values. We provide four extensions to the bicriterion approach for anticlustering: First, we present a novel exact algorithm that returns multiple globally optimal partitions with regard to the dispersion. Despite its theoretical computational hardness, our method scales to rather large data sets (e.g., $N > 5000$ for $K = 3$ groups). Second, by extending existing optimal and heuristic approaches (including BILS), we evaluate how to best obtain a high diversity on top of an optimal dispersion provided by our algorithm. Third, we allow that the dispersion and the diversity can be optimized on the basis of different dissimilarity matrices. Forth, by reformulating the k-means clustering criterion as an average diversity criterion, we allow that other anticlustering objectives of between-group similarity can be optimized alongside the dispersion instead of only the diversity. The original BILS as well as our extensions are available via the free and open source R package `anticlust`. Practical examples are included to illustrate their application.
